EGGs are adhesive!

### 2025-03-14 calco/main.tex
minima

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many of these general results could be recast and uniformly proved once and for all.

@@

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the classical results from the theory of graph 
transformation can still be obtained~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being adhesive can be a daunting task. For this reason,  sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
	Then it holds:
	\begin{enumerate}
		\item the right face of the cube is a pullback too;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

Having prove stability of $\pbc$-pushouts, we now turn to prove that they are Van Kampen with respect to regular monos.
%
A proof is found in \Cref{proof:pbvk}.

@@

\noindent
\parbox{11.5cm}{\begin{restatable}{lemma}{epo}\label{lem:epo}
	Suppose that the square on the right is a pushout in $\EqHyp$ with $m$ an arrow in $\pbc$. If $\mathcal{G}_1$, $\mathcal{G}_2$ and $\mathcal{G}_3$ are e-hypergraphs, then $\mathcal{P}$ is an object of $\egg$ too.
\end{restatable}}\hfill 
\parbox{4cm}{\xymatrix@R=15pt{\mathcal{G}_1 \ar[r]^{h}\ar[d]_{m}&\mathcal{G}_2\ar[d]^{n}\\\mathcal{G}_3\ar[r]_{z}&\mathcal{P}}}

@@

We turn now to the labelled context.

\begin{definition} We say that an object $(\mathcal{H}, l)$ of $\EqHyp_\Sigma$ is a \emph{labelled e-hypergraph} if $\mathcal{H}$ is an e-hypergraph. We define the category $\egg_\Sigma$ as the full subcategory of $\EqHyp_\Sigma$ given by labelled e-hypergraphs. We will denote by $j_\Sigma$ the corresponding inclusion functor .
\end{definition}

@@

\begin{remark}\label{rem:algegg}
	Given a signature $\Sigma = (O_\Sigma, \ari_\Sigma)$,
	the hypergraph 
	$R(\mathcal{G}^\Sigma) = (O_\Sigma, 1, 1, \ari_\Sigma, \gamma_1, \id{1})$
	is an object of $\egg$.
	Indeed, under the identification of $1^\star$ with $\mathbb{N}$, the kernel $(S, \pi_1, \pi_2)$  of $\id{\mathbb{N}} \circ \ari_\Sigma$, is given by
	\[S:=\{(o_1, o_2)\in O_\Sigma \times O_\Sigma \mid \ari_\Sigma(o_1)=\ari_\Sigma(o_2)\}\]
	equipped with the two projections.  Onthe other hand $\id{\mathbb{N}} \circ \gamma_1 \circ \pi_1$ and $\id{\mathbb{N}}\circ \gamma_1 \circ \pi_2$ are both the function $O_\Sigma \to \mathbb{N}$ constant in $1$.
\end{remark}

\Cref{rem:algegg} now implies at once the following result.
\begin{proposition}\label{prop:varie}
	Let $\Sigma$ be an algebraic signature. Then the following hold true:
	\begin{enumerate}
		\item $\egg_\Sigma$ is equivalent to $\egg/R(\mathcal{G}^\Sigma)$.
		\item There exists a functor $W_\Sigma \colon \egg_\Sigma\to \Sigma$ forgetting the labeling which creates all colimits, pullbacks and equalizers. 
	\end{enumerate}
\end{proposition}

Let $\pbe_\Sigma$ be the class of morphisms in $\egg_\Sigma$ whose image in $\egg$ lies in $\pbe$. Notice that $\pbe_\Sigma$ is also the class of arrows whose image through $j_\Sigma$ is in $\pbc_\Sigma$.

By \Cref{thm:slice-functors} we get the following.

@@

\begin{definition}\label[definintion]{def:GEqTGs}
	Given a signature $\Sigma$,  we say that an object $(\mathcal{H}, l)$ of $\EqTG_{\Sigma}$ is a \emph{e-term graph} if $\mathcal{H}$ is an e-hypergraph. We define the category $\eg$ as the full subcategory of $\EqTG_{\Sigma}$ given by e-term graphs and denote by $k_\Sigma$ the corresponding inclusion.
\end{definition}

\begin{remark}
	By definition we also have an inclusion functor $K_\Sigma\colon \eg\to \egg_\Sigma$.
\end{remark}

\noindent 
\parbox{7.5cm}{\begin{remark}
	Now that we have introduced all the characters of this work, it is worth to give a visual map of all the categories that we have introduced in this paper and of the relationships between them. We will use the curved arrows to denote full and faithful inclusions.
\end{remark}}\hfill \parbox{4cm}{\xymatrix@R=15pt@C=15pt{\eg \ar@{^{(}->}[r]^-{K_\Sigma} \ar@{^{(}->}[d]_{k_\Sigma} & \egg_\Sigma \ar@{^{(}->}[d]^{j_\Sigma} \ar[r]^{W_\Sigma}& \egg \ar@{^{(}->}[d]^{I}\\ \EqTG_{\Sigma} \ar@{^{(}->}[r]_{J_\Sigma}\ar[d]_{S_\Sigma}& \EqHyp_\Sigma \ar[r]_{V_\Sigma} \ar[d]_{T_\Sigma}& \EqHyp \ar[d]_{T}\\ \tg \ar@{^{(}->}[r]_{I_\Sigma}& \hyp_{\Sigma} \ar[r]_{U_\Sigma}  &\hyp}}

We can now prove our last three results. The interested reader can find the proof in \Cref{proof:tlim}.

\begin{restatable}{proposition}{limt}\label{prop:limt}
	$\eg$ has equalizer, binary products and pullbacks and they are created by $k_\Sigma$.
\end{restatable}

Let now $\mathcal{T}_\Sigma$ be the class of morphisms of $\eg$ which are sent by $k_\Sigma$ to the class $\mathcal{T}$. Then \Cref{colim,lem:epo,prop:po} allow us to deduce the following result.

\begin{restatable}{proposition}{pot}\label{prop:pot}
	$\eg$ has $\mathcal{T}_\Sigma$-pushouts which are created by $k_\Sigma$.
\end{restatable}

@@

\section{Pros and cons of adhesive rewriting}
\label{rewriting}
The previous sections has shown how term graphs with equivalence,
possibly closed with respect to operator composition,
can be described as a suitable $\mathcal{M}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Being a purely graphical representation, EGGS allows for redundancy and possibly ambiguity,
so that e.g. a term such as $a / a$ admits a few different representations as a term graph, as shown below
%
%\[a \div a \mbox{ con 2 a sharate, con 2 a nella stessa classe di equivalenze, e con 2 a separate}\]

@@

Admittedly, the third one is illegal, i.e. it is an object of $\EqTGs$ but not one of $\GEqTGs$: constants have no input, 

@@

$\mathcal{G}  \downarrow \GEqTGs$. 

@@

The same construction carries on when restricting to the full category of $\GEqTGs$ of those objects whose 

@@

What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach.
% to the category at hand. 
However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, thus in $\pbc_\Sigma$,
while the second component may not belong to $\pbc_\Sigma$.

@@

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of $\mathcal{M}$-adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited and generalised the notions of hyper-graph and term graph with equivalence, and
we extended it in order to capture EGGS as term graphs satisfying a suitable closure property.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\GEqTGs$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
This fact allows for exploiting the properties 
of the $\mathcal{M}$-adhesive framework when discussing about the optimisation process.

\emph{Future works.}
Our result on $\GEqTGs$ opens a few threads of research. The first is to check how the $\mathcal{M}$-adhesivity 
of EGGs can be pushed to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be spans whose left-hand side is an identity and 
right-hand side is a regular mono (possibly with negative application 
conditions), and as such they fit the mould of rewriting on left-linear rewriting in $\mathcal{M}$-adhesive categories. 
%The consequences for usability on the restriction of the matches to arrows that are injective on equivalence classes
%though need to be further investigated.
%
Also, it still needs to be exploited how parallelism and causality, the key features for DPO rewriting
on $\mathcal{M}$-adhesive categories, can be exploited in the context of EGGs implementation. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hypergraphs can be generalised to hierarchical hypergraphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges~\cite{ghicaZan,CastelnovoGM24}, which is useful for adding structural information, such as encapsulation and sandboxing.

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents (acyclic) term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly equivalent to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework.

@@

 		\item  We begin by computing to get:
		\[g \circ h \circ \pi_f^1 =  t \circ f \circ \pi_f^1     =  t \circ f \circ \pi_f^2     =  g \circ h \circ \pi_f^2\]
		so that  existence  and uniqueness of the wanted $k_h$ follow at once from the the universal property of $K_g$ as the pullback of $g$ along itself.

@@

\noindent
\parbox{11cm}{
\begin{remark}\label{rem:disj}
	Notice that extensivity entails that coproducts are \emph{disjoint}, i.e.~ that the pullback between two coproduct injections is given by the initial object (with initial maps ad coprojections). To see this, let $\{X, \{j_i\}_{i\in I}\}$ be the coproduct of the family $\{X_i\}_{i\in I}$ and let $k$ be an element of $I$. Then  $(X_k, \{t_i\}_{i\in I})$ with
	\[t_i=\begin{cases}
		\id{X_k} & i=k\\?_{X_k} & i\neq k
	\end{cases}\]
	is a coproduct and so the squares on the right are pullbacks.
\end{remark}}\hfill\parbox{2cm}{\vspace{-1.5cm}\xymatrix@R=15pt{\emptyset \ar@{>->}[d]_{?_{X_i}} \ar@{>->}[r]_{?_{X_k}}& X_k \ar@{>->}[d]^{j_k} \\ X_i \ar@{>->}[r]^{j_i} & X} \xymatrix@R=15pt{X_k \ar[r]_{\id{X}} \ar[d]_{\id{X}}& X_k \ar@{>->}[d]^{j_k}\\X_k \ar@{>->}[r]^{j_k} & X}}

@@

\begin{proof}\label{proof:kerset}
	We can start noticing that $m$, being the pullback of $n$ is mono too. This, in turn, entails that $m'$, being the pullback of $m$, is mono, and so $n'$ is injective too because in $\Set$, as in any adhesive category, monomorphisms are stable under pushouts.

@@

@@

We focus now on $K_3$. Computing we get
	\begin{align*}
	&g\circ \iota \circ w\circ p_3=g\circ c\circ \iota'\circ p_3=d\circ g'\circ \iota'\circ p_3=d\circ \pi^1_d\circ j_3\\=&d\circ \pi^2_d\circ j_3=d\circ g'\circ \iota' \circ q_3=g\circ c\circ \iota'\circ q_3=g\circ \iota \circ w\circ q_3
	\end{align*}

\noindent 
\parbox{3cm}{\xymatrix@R=15pt{ K_3 \ar[r]^{p_3} \ar@{.>}[dr]^{\phi_3} \ar[d]_{q_3}& E' \ar[dr]^{\iota'} \\E' \ar[dr]_{\iota'}&K_c \ar[r]^{\pi^1_c} \ar[d]_{\pi^2_c} & C'  \ar[d]^{c}\\ & C' \ar[r]_{c} & C }}\hfill\parbox{9.5cm}{By \Cref{lem:po_set} $g\circ \iota$ is a monomorphism, thus $w\circ p_3=w\circ q_3$ and therefore:
\[c\circ \iota'\circ p_3=\iota\circ w\circ p_3=\iota \circ w\circ q_3=c\circ \iota'\circ q_3\]
Thus we get the dotted $\phi_3\colon K_3\to K_c$ in the diagram aside.}

@@

 	\smallskip
 \parbox{7.5cm}{We can go further. The outer part of the diagram on the right commutes, so that we have the dotted $\psi_0\circ K_b\to K_0$.
 Now, on the one hand $\phi_0\circ \psi_0=\id{K_b}$ because if we compute we get:
  \[(\pi^1_b, \pi^2_b)\circ \phi_0\circ \psi_0=(p_0, q_0)\circ \psi_0=(\pi^1_b, \pi^2_b)\]}\hfill \parbox{4cm}{\vspace{-.5cm}\xymatrix{K_b \ar[dr]_{(\pi^1_b, \pi^2_b)}  \ar@{>->}@/^.4cm/[rr]^{k_{n'}} \ar@{.>}[r]_{\psi_0}& K_0 \ar@{>->}[r]_{j_0}  \ar[d]^{(p_0, q_0)}& K_d   \ar[d]^{(\pi^1_d, \pi^2_d)} \\
 	& B'\times B' \ar@{>->}[r]_-{n'\times n'} & D'\times D' }
 	}

@@

 On the other hand, 
 \[\theta \circ k_{g'}\circ x_0=\theta \circ k_{n'}\circ h_0=\gamma \circ h_0=\gamma \circ k_{f'}\circ h'_0=\delta \circ k_{m'}\circ h'_0=\delta \circ x_0 \]
 and we can conclude that $\theta \circ k_{g'}=\delta$ as wanted.

@@

This section is devoted to recall some basic facts about the so called \emph{slice categories}.
\begin{definition}\index{category!slice -}
	Let $X$ be an object of a category $\X$, we will define the following two categories.

>>>>

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many of these general results can be recast and uniformly proved once and for all.

@@

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefiots of adhesivity
%the classical results from the theory of graph transformation 
can still be obtained~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
	Then it holds
	\begin{enumerate}
		\item the right face of the cube is a pullback too;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

Having proved stability of $\pbc$-pushouts, we now turn to prove that they are Van Kampen with respect to regular monos.
%
A proof is found in \Cref{proof:pbvk}.

@@

\noindent
\parbox{11.5cm}{\begin{restatable}{lemma}{epo}\label{lem:epo}
	Suppose that the square on the right is a pushout in $\EqHyp$ with $m$ an arrow in $\pbc$. If $\mathcal{G}_1$, $\mathcal{G}_2$ and $\mathcal{G}_3$ are e-hypergraphs, then $\mathcal{P}$ is so too.
	%an object of $\egg$ too.
\end{restatable}}\hfill 
\parbox{4cm}{\xymatrix@R=15pt{\mathcal{G}_1 \ar[r]^{h}\ar@{>->}[d]_{m}&\mathcal{G}_2\ar[d]^{n}\\\mathcal{G}_3\ar[r]_{z}&\mathcal{P}}}

@@

We now turn our attention to the labelled context.

\begin{definition} We say that an object $(\mathcal{H}, l)$ of $\EqHyp_\Sigma$ is a \emph{labelled e-hypergraph} if $\mathcal{H}$ is an e-hypergraph. We define the category $\egg_\Sigma$ as the full subcategory of $\EqHyp_\Sigma$ given by labelled e-hypergraphs. We denote by $J_\Sigma$ the corresponding inclusion functor .
\end{definition}

@@

\begin{remark}\label{rem:algegg}
	Given a signature $\Sigma = (O_\Sigma, \ari_\Sigma)$,
	the hypergraph 
	$R(\mathcal{G}^\Sigma) = (O_\Sigma, 1, 1, \ari_\Sigma, \gamma_1, \id{1})$
	is an object of $\egg$.
	Indeed, under the identification of $1^\star$ with $\mathbb{N}$, the kernel $(S, \pi_1, \pi_2)$  of $\id{\mathbb{N}} \circ \ari_\Sigma$, is given by
	$S:=\{(o_1, o_2)\in O_\Sigma \times O_\Sigma \mid \ari_\Sigma(o_1)=\ari_\Sigma(o_2)\}$
	equipped with the two projections.  On the other hand $\id{\mathbb{N}} \circ \gamma_1 \circ \pi_1$ and $\id{\mathbb{N}}\circ \gamma_1 \circ \pi_2$ are both the function $O_\Sigma \to \mathbb{N}$ constant in $1$.
\end{remark}

\Cref{rem:algegg} now implies at once the following result.
\begin{proposition}\label{prop:varie}
	Let $\Sigma$ be an algebraic signature. Then it holds
	\begin{enumerate}
		\item $\egg_\Sigma$ is equivalent to $\egg/R(\mathcal{G}^\Sigma)$.
		\item There exists a functor $W_\Sigma \colon \egg_\Sigma\to \Sigma$ forgetting the labeling which creates all colimits, pullbacks and equalizers. 
	\end{enumerate}
\end{proposition}

Let $\pbe_\Sigma$ be the class of morphisms in $\egg_\Sigma$ whose image in $\egg$ lies in $\pbe$. Notice that $\pbe_\Sigma$ is also the class of arrows whose image through $J_\Sigma$ is in $\pbc_\Sigma$.
%
By \Cref{thm:slice-functors} we get the following result.

@@

\begin{definition}\label[definintion]{def:GEqTGs}
	Given a signature $\Sigma$,  we say that an object $(\mathcal{H}, l)$ of $\EqTG_{\Sigma}$ is a \emph{e-term graph} if $\mathcal{H}$ is an e-hypergraph. We define the category $\eg$ as the full subcategory of $\EqTG_{\Sigma}$ given by e-term graphs and denote by $K_\Sigma$ the corresponding inclusion.
\end{definition}

\begin{remark}
	By definition we also have an inclusion functor $Y_\Sigma\colon \eg\to \egg_\Sigma$.
\end{remark}

\noindent 
\parbox{7.5cm}{\begin{remark}
	Now that we have introduced all the of this work in place, it is worthwhile to give a visual map of all the categories that we have introduced in this paper and of the relationships between them. We will use the curved arrows to denote full and faithful inclusions.
\end{remark}}\hfill \parbox{4cm}{\xymatrix@R=15pt@C=15pt{\eg \ar@{^{(}->}[r]^-{Y_\Sigma} \ar@{^{(}->}[d]_{K_\Sigma} & \egg_\Sigma \ar@{^{(}->}[d]^{j_\Sigma} \ar[r]^{W_\Sigma}& \egg \ar@{^{(}->}[d]^{I}\\ \EqTG_{\Sigma} \ar@{^{(}->}[r]_{J_\Sigma}\ar[d]_{S_\Sigma}& \EqHyp_\Sigma \ar[r]_{V_\Sigma} \ar[d]_{T_\Sigma}& \EqHyp \ar[d]_{T}\\ \tg \ar@{^{(}->}[r]_{I_\Sigma}& \hyp_{\Sigma} \ar[r]_{U_\Sigma}  &\hyp}}

We can now prove our last three results. The reader can find the proof in \Cref{proof:tlim}.

\begin{restatable}{proposition}{limt}\label{prop:limt}
	$\eg$ has equalizers, binary products and pullbacks and they are created by $K_\Sigma$.
\end{restatable}

Let now $\mathcal{T}_\Sigma$ be the class of morphisms of $\eg$ which are sent by $K_\Sigma$ to the class $\mathcal{T}$. Then \Cref{colim,lem:epo,prop:po} allow us to deduce the following result.

\begin{restatable}{proposition}{pot}\label{prop:pot}
	$\eg$ has $\mathcal{T}_\Sigma$-pushouts which are created by $K_\Sigma$.
\end{restatable}

@@

\section{Pros and cons of adhesive rewriting}
\label{rewriting}
The previous sections have shown how term graphs with equivalence,
possibly closed with respect to operator composition,
can be described as a suitable $\mathcal{M}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Being a purely graphical representation, EGGS allows for redundancy and possibly ambiguity,
so that e.g. a term such as $a / a$ admits a few different representations as a term graph, as shown below
%
%\[a \div a \mbox{ con 2 a sharate, con 2 a nella stessa classe di equivalenze, e con 2 a separate}\]

@@

Admittedly, the third one is illegal, i.e. it is an object of $\EqTGs$ but not one of $\eg$:
%$\GEqTGs$: 
constants have no input, 

@@

$\mathcal{G}  \downarrow \eg$.
%\GEqTGs$. 

@@

The same construction carries on when restricting to the full subcategory of $\eg$
%$\GEqTGs$ 
of those objects whose 

@@

\newpage
What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach.
% to the category at hand. 
However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, thus in $\pbc_\Sigma$,
while the second component may not belong to $\pbc_\Sigma$.

@@

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of $\mathcal{M}$-adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited the notion of hyper-graph and introduced  term graph with equivalence, 
proving that they are $\mathcal{M}$-adhesive categories, and we extended this resulr in order to capture
also prove the same property for EGGS as term graphs with equivalence satisfying a suitable closure constraint.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\eg$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
This fact allows for exploiting the properties 
of the $\mathcal{M}$-adhesive framework when discussing about the optimisation process.

\emph{Future works.}
Our result on $\eg$ opens a few threads of research. The first is to check how the $\mathcal{M}$-adhesivity 
of EGGs can be pushed to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be spans whose left-hand side is an identity and 
right-hand side is a regular mono (possibly with negative application 
conditions), and as such they fit the mould of rewriting on left-linear rewriting in $\mathcal{M}$-adhesive categories. 
%The consequences for usability on the restriction of the matches to arrows that are injective on equivalence classes
%though need to be further investigated.
%
Also, it still needs to be exploited how parallelism and causality, the key features for DPO rewriting
on $\mathcal{M}$-adhesive categories, can be exploited in the context of implementing the 
EGGs updates. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hypergraphs can be generalised to hierarchical hypergraphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges~\cite{ghicaZan,CastelnovoGM24}, which is useful for adding structural information, such as encapsulation and sandboxing.

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents (acyclic) term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly equivalent to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework
as they hold for standard graph transformation.

@@

 		\item  We begin by computing 
		\[g \circ h \circ \pi_f^1 =  t \circ f \circ \pi_f^1     =  t \circ f \circ \pi_f^2     =  g \circ h \circ \pi_f^2\]
		so that  existence  and uniqueness of the wanted $k_h$ follow at once from the the universal property of $K_g$ as the pullback of $g$ along itself.

@@

\noindent
\parbox{11cm}{
\begin{remark}\label{rem:disj}
	Notice that extensivity entails that coproducts are \emph{disjoint}, i.e. that the pullback between two coproduct injections is given by the initial object (with initial maps ad coprojections). To see this, let $\{X, \{j_i\}_{i\in I}\}$ be the coproduct of the family $\{X_i\}_{i\in I}$ and let $k$ be an element of $I$. Then  $(X_k, \{t_i\}_{i\in I})$ with
	\[t_i=\begin{cases}
		\id{X_k} & i=k\\?_{X_k} & i\neq k
	\end{cases}\]
	is a coproduct and so the squares on the right are pullbacks.
\end{remark}}\hfill\parbox{2cm}{\vspace{-1.5cm}\xymatrix@R=15pt{\emptyset \ar@{>->}[d]_{?_{X_i}} \ar@{>->}[r]_{?_{X_k}}& X_k \ar@{>->}[d]^{j_k} \\ X_i \ar@{>->}[r]^{j_i} & X} \xymatrix@R=15pt{X_k \ar[r]_{\id{X}} \ar[d]_{\id{X}}& X_k \ar@{>->}[d]^{j_k}\\X_k \ar@{>->}[r]^{j_k} & X}}

@@

\begin{proof}\label{proof:kerset}
	We can start noticing that $m$, being the pullback of $n$ is mono too. This, in turn, entails that $m'$, being the pullback of $m$, is mono, and so $n'$ is injective too because in $\Set$, as in any adhesive category, monomorphisms are stable under pushouts.  
\vspace{.1cm}

@@

\vspace{.1cm}

@@

We focus now on $K_3$. By computing we get
	\begin{align*}
	&g\circ \iota \circ w\circ p_3=g\circ c\circ \iota'\circ p_3=d\circ g'\circ \iota'\circ p_3=d\circ \pi^1_d\circ j_3\\=&d\circ \pi^2_d\circ j_3=d\circ g'\circ \iota' \circ q_3=g\circ c\circ \iota'\circ q_3=g\circ \iota \circ w\circ q_3
	\end{align*}

\noindent 
\parbox{3cm}{\xymatrix@R=15pt{ K_3 \ar[r]^{p_3} \ar@{.>}[dr]^{\phi_3} \ar[d]_{q_3}& E' \ar[dr]^{\iota'} \\E' \ar[dr]_{\iota'}&K_c \ar[r]^{\pi^1_c} \ar[d]_{\pi^2_c} & C'  \ar[d]^{c}\\ & C' \ar[r]_{c} & C }}\hfill\parbox{9.5cm}{By \Cref{lem:po_set} $g\circ \iota$ is a monomorphism, thus $w\circ p_3=w\circ q_3$ and therefore
\[c\circ \iota'\circ p_3=\iota\circ w\circ p_3=\iota \circ w\circ q_3=c\circ \iota'\circ q_3\]
Thus we get the dotted $\phi_3\colon K_3\to K_c$ in the diagram aside.}

@@

 	\smallskip
 \parbox{7.5cm}{We can go further. The outer part of the diagram on the right commutes, so that we have the dotted $\psi_0\circ K_b\to K_0$.
 Now, on the one hand $\phi_0\circ \psi_0=\id{K_b}$ because if we compute we get
  \[(\pi^1_b, \pi^2_b)\circ \phi_0\circ \psi_0=(p_0, q_0)\circ \psi_0=(\pi^1_b, \pi^2_b)\]}\hfill \parbox{4cm}{\vspace{-.5cm}\xymatrix{K_b \ar[dr]_{(\pi^1_b, \pi^2_b)}  \ar@{>->}@/^.4cm/[rr]^{k_{n'}} \ar@{.>}[r]_{\psi_0}& K_0 \ar@{>->}[r]_{j_0}  \ar[d]^{(p_0, q_0)}& K_d   \ar[d]^{(\pi^1_d, \pi^2_d)} \\
 	& B'\times B' \ar@{>->}[r]_-{n'\times n'} & D'\times D' }
 	}

@@

 On the other hand
 \[\theta \circ k_{g'}\circ x_0=\theta \circ k_{n'}\circ h_0=\gamma \circ h_0=\gamma \circ k_{f'}\circ h'_0=\delta \circ k_{m'}\circ h'_0=\delta \circ x_0 \]
 and we can conclude that $\theta \circ k_{g'}=\delta$ as wanted.

@@

This section is devoted to recall some basic facts about the so called \emph{slice categories}.
\begin{definition}\index{category!slice -}
	Let $X$ be an object of a category $\X$. We define the following two categories.
